Modeling and Simulation of Load Balancing Strategies for Computing in High Energy Physics

### 2018-10-22 paper/paper.tex
Some linguistic improvements

\abstract{%
	The amount of data to be processed by experiments in high energy physics (HEP) is tremendously increasing in the coming years. To cope with this increasing load, most efficient usage of the resources is mandatory. Furthermore, the computing resources for user jobs in HEP will be increasingly distributed and heterogeneous, resulting in more difficult scheduling due to the increasing complexity of the system.
	We aim to create a simulation for the WLCG helping the HEP community to solve both challenges: a more efficient utilization of the grid and coping with the rising complexity of the system.
	There is currently no simulation existing helping the operators of the grid to make the correct decisions while optimizing the load balancing strategy.

@@

	Finally, the simulation results are validated using real word performance data. Our results suggest that simulating larger parts of the grid is feasible and can help to optimize the utilization of the grid.
}%
\maketitle
%

@@

We aim to create a simulation for the WLCG helping the HEP community to solve both challenges: a more efficient utilization of the grid and coping with the rising complexity of the system.
Currently, there is no model or simulation in existence to evaluate the load balancing strategies of the grid, helping the operators to make the correct decisions while optimizing the load balancing strategy.
\Cref{related} describes different attempts to simulate the WLCG or parts of it. However, they showed scalability issues or the simulations do not have the necessary detail level to simulate effects on working nodes that is necessary to evaluate load balancing strategies.

Optimizing load balancing strategies without the help of simulations requires stable load, in order to preserve the comparability of the different strategies. This is neither possible to achieve for the WLCG nor for a single center like GridKa, because the nature of the load is highly dynamical. Another option would be to use an isolated test system for the evaluation. However, the system under inspection is too large and thus a similar test systems would not be financially affordable. A smaller test system could not deliver correct results because too much of the complexity of the system would be taken away.
These reasons lead to the conclusion that a model and simulation of the performance of computing jobs is required to evaluate different load balancing strategies.

@@

The contribution of this paper is the modeling and simulating of CMS computing jobs at the Tier 1 center GridKa as a proof of concept.
The model and simulation are created using the Palladio simulator, see \Cref{sec:palladio} for details on Palladio. We chose Palladio to achieve a scalable simulation because it is a comparably abstract simulator (\Cref{related} gives an overview on other simulators and approaches).
This leads to the research question how we can use Palladio to evaluate load balancing strategies for the CMS computing jobs, because the simulation of load balancing strategies is a new use case for Palladio. To realise it we created an extension for the simulator that enables us to simulate load balancing strategies and furthermore, we created an automated model parameter analysis and model creation, in order to increase the usability of the simulation.
This model makes it possible to evaluate different load balancing strategies for GridKa and enables us to draw conclusions about the best optimizations of the load balancing for the WLCG. Also, it can be used as a foundation for models and simulations of larger parts of the WLCG.

This paper is organized as follows: \Cref{sec:palladio} presents the Palladio simulator as a foundation of the simulation. \Cref{sec:process} gives a overview on the simulation process. \Cref{sec:model} describes how different elements of the computing jobs and the computing resources are represented in our model. \Cref{sec:param} explains how we determine the model parameters and the automated model creation works. The validation is presented in \Cref{validation} and \Cref{related} puts our work in context with the related work. In \Cref{outlook} possible design decisions and how they could be modeled and simulated are presented. Finally, \Cref{conlusion} gives a conclusion on this paper.

@@

\section{Simulation Process}
\label{sec:process}
To receive a valid model of a site and its jobs, we created the simulation process depicted in \Cref{process}. This process consists of five phases: Matching of data sets, grouping the data for job types, analysing the model parameters, generating a blueprint model, and lastly injecting the parameters into this model. Note, only a part of the data is used to calibrate the model, the other part is used for its validation.

The starting point of the simulation process is matching the data sets from different data sources as a prerequisite  of the analysis. In our proof of concept we have three data sources: The CERN WMArchive, CMS GridKa job data and GridKa node performance data. These data sets have to be matched because each set contains different information about an executed job, but none of them contains the complete information about it.

@@

The next step is the generation of the model of GridKa and the injection of the previously extracted performance parameters. Both steps do not require manually modeling. At this point the users have a valid and calibrated model of GridKa with its resource and the workload. The properties of this model are described in \Cref{sec:model}.

@@

\Cref{seff} shows the behaviour of the job during the simulation. After a job is scheduled on a resource container, it acquires the specified number of job slots or waits if the node has not sufficient free slots. Afterwards, the CPU and I/O demands of the job are processed. To avoid one big I/O operation at the start of a job and to make the simulated processing of I/O more continuous, it is modeled using iterations that reflect the number of processed events. Thus, for each iteration there are I/O and CPU resources needed. Rather than simulating each event, we simulate a fraction of the processed events, because the simulation of each event would harm the scalability of the simulation. 
While processing, the job uses the specified number of threads to process the CPU demand. After the job finished, it releases its acquired job slots. All of these parameters are specified as a probability function and are extracted from measured performance data as described in \Cref{sec:param}.

@@

At GridKa every type of computing job can run on every node, thus each job type is allocated on each computing node in the allocation model.

@@

We extended Palladio with a \textit{load balancing strategy action}\footurl{https://doi.org/10.5281/zenodo.1464569}, so that we can simulate load balancing strategies.
This load balancing strategy action is contained in a load balancing component.
In the assembly model all job components are connected with the load balancing component, which is the entry point of the system.
This action decides on which node a job should be scheduled based on the currently free job slots of each working node. The load balancing strategies can be extended to evaluate different strategies.

@@

To obtain a comprehensive view both of the resource usage of CMS jobs executed at a WLCG grid site as well as the resources available there to process them, combining information from multiple data sources is required:
\begin{itemize}
    \item We acquire historical job monitoring information from several global CMS monitoring systems: Job reports from JobMonitoring and Workflow and Data Management framework job reports (FWJR documents) are archived on the \emph{analytix} Hadoop cluster.
    The latter are being archived on the cluster as part of the WMArchive project \cite{WMArchiveWorkloadArchive2018}.
    We collect job information using the \emph{CMSSpark} framework by Kuznetsov et al. \cite{cmsspark}.

@@

\begin{figure}[h]
	\centering
	\vspace{-10px}
	\begin{subfigure}{0.48\linewidth}
		\includegraphics[width=\linewidth]{images/validation/throughputs}
	\end{subfigure}
	\begin{subfigure}{0.48\linewidth}
		\includegraphics[width=\linewidth]{images/validation/utilizations}
	\end{subfigure}
	\vspace{-10px}
	\caption{Measured and simulated throughputs and CPU efficiency, job data from May 2018}
	\label{fig:validation}
\end{figure}
\vspace{10px}
\noindent To validate the model and simulation we simulated CMS jobs at GridKa in May 2018.
The job performance data was split into two equal sized calibration and test datasets by randomly sampling reports from the full data set.
For the validation we considered the following metrics: Overall throughput of jobs, shares of the job types, and the overall CPU efficiency of the site.
These metrics describe how well the simulation matches the measured performance data.
Comparisons between simulation results and measurements from the system are shown in \Cref{fig:validation}.
One result is that simulated throughput matches the reference with a difference of about 5.6\%.
Furthermore, the shares of the simulated and measured job types also match accurately.

@@

Simulating one week of CMS jobs at GridKa takes about 24 minutes on a laptop.
Thus, it is feasible to simulate several sites at once in a reasonable time, however simulating larger parts of the grid is future work.

@@

There are a variety of grid and cloud simulators existing, of which SimGrid and CloudSim are two popular ones. 
However, both of them are implemented as libraries and not as standalone simulators. Thus, both are missing graphical model editors which makes the modeling of the computing jobs time-consuming \cite{simgrid,cloudsim}. The support of graphical editors is a big advantage of Palladio and increases its usability and efficiency. Another advantage of Palladio is the separation of the system into different models, which enables one to change on part of the system without modifying others. Furthermore, both do lack the support of an integrated visualization tool to display the results.

CACTOS is an approach to cloud infrastructure automation and optimization \cite{cactos}.
One part of the approach is the CACTOS Runtime Toolkit for monitoring and resource management and another part is the CACTOS Prediction Toolkit for evaluation of alternative data center deployment scenarios and resource management algorithms. The Prediction Toolkit is built on Palladio and SimuLizar.
A first case studies in the context of chemical computing showed that the Prediction Toolkit has a high accuracy, but it was not tested on large systems yet \cite{rapidtesting}. However, due the required Runtime Toolkit CACTOS was not applicable in our use case.

@@

One possible design decision is a more sophisticated load balancing strategy that aims for a better resource utilization of the working nodes by scheduling a balanced share of CPU and I/O heavy jobs to it. To simulate this a new scheduling strategy must be implemented.

Another design decision would be to experiment with different sizes of pilot jobs, assuming that smaller pilot jobs result in a better utilization due less wasted CPU capacity in the draining phase of a pilot. For this use case the model must be extended with the explicit modeling of pilots jobs.

Additionally, the effects of integrating cloud resources could be simulated. The model must be extended with the cloud resources and the required network load to access the files from a site. Then a good load balancing strategy for the  integration of cloud resources could be evaluated, like only scheduling CPU heavy jobs to cloud resources and keeping I/O heavy jobs near the data.

Also the effects of data caches could be simulated. Firstly, several sites must be modeled to simulate their interactions. Secondly, the jobs must be parametrized with the possibility of remote access. Afterwards, the data cache with its hit rate should be modeled. Finally, the performance improvements of the data cache by transferring data from the cache and not from sites can be simulated.
\section{Conclusion}
\label{conlusion}

@@

We successfully simulated the CMS computing jobs at GridKa. To enable easy usage of the simulation we implemented an automated model parameter extraction and model creation. Furthermore, we extended Palladio to be able to simulate load balancing strategies. Our first results are promising and suggest that this approach can be used to simulate load balancing strategies accurately. Regarding scalability we are confident that several computing centers can be simulated in a sufficient time, however this is part of future work.

>>>>

\abstract{%
	The amount of data to be processed by experiments in high energy physics (HEP) will increase tremendously in the coming years. To cope with this increasing load, most efficient usage of the resources is mandatory. Furthermore, the computing resources for user jobs in HEP will be increasingly distributed and heterogeneous, resulting in more difficult scheduling due to the increasing complexity of the system.
	We aim to create a simulation for the WLCG helping the HEP community to solve both challenges: a more efficient utilization of the grid and coping with the rising complexity of the system.
	There is currently no simulation in existence which helps the operators of the grid to make the correct decisions while optimizing the load balancing strategy.

@@

	Finally, the simulation results are validated using real-word performance data. Our results suggest that simulating larger parts of the grid is feasible and can help to optimize the utilization of the grid.
}%
\maketitle
%

@@

We aim to create a simulation for the WLCG helping the HEP community to solve both challenges: a more efficient utilization of the grid and coping with the rising complexity of the system.
Currently, there is no model or simulation in existence to evaluate the load balancing strategies of the grid, helping the operators to make the correct decisions while optimizing the load balancing strategy.
\Cref{related} describes different attempts to simulate the WLCG or parts of it. However, they either showed scalability issues or the simulations do not have the necessary detail level to simulate effects on working nodes that is necessary to evaluate load balancing strategies.

Optimizing load balancing strategies without the help of simulations requires stable load in order to preserve the comparability of the different strategies. This is neither possible to achieve for the WLCG nor for a single center like GridKa, because the nature of the load is highly dynamical. Another option would be to use an isolated test system for the evaluation. However, the system under inspection is too large and thus a similar test system would not be financially affordable. A smaller test system could not deliver the correct results because too much of the complexity of the system would be taken away.
These reasons lead to the conclusion that a model and simulation of the performance of computing jobs is required to evaluate different load balancing strategies.

@@

The contribution of this paper is the modeling and simulating of CMS computing jobs at the Tier 1 center GridKa as a proof of concept.
The model and simulation are created using the Palladio simulator, see \Cref{sec:palladio} for details on Palladio. We chose Palladio to achieve a scalable simulation because it is a comparably abstract simulator (\Cref{related} gives an overview on other simulators and approaches).
This leads to the research question of how we can use Palladio to evaluate load balancing strategies for the CMS computing jobs, because the simulation of load balancing strategies is a new use case for Palladio. To answer this question, we created an extension for the simulator that enables us to simulate load balancing strategies. Furthermore, we created an automated model parameter analysis and model creation in order to increase the usability of the simulation.
This model makes it possible to evaluate different load balancing strategies for GridKa and enables us to draw conclusions about the best optimizations of the load balancing for the WLCG. Also, it can be used as a foundation for models and simulations of larger parts of the WLCG.

This paper is organized as follows: \Cref{sec:palladio} presents the Palladio simulator as a foundation of the simulation. \Cref{sec:process} gives a overview on the simulation process. \Cref{sec:model} describes how different elements of the computing jobs and the computing resources are represented in our model. \Cref{sec:param} explains how we determine the model parameters and the automated model creation works. The validation is presented in \Cref{validation} and \Cref{related} puts our work into context with the related work. In \Cref{outlook}, possible design decisions and how they could be modeled and simulated, are presented. Finally, \Cref{conlusion} gives a conclusion on this paper.

@@

\section{Simulation Process}
\label{sec:process}
To receive a valid model of a site and its jobs, we created the simulation process depicted in \Cref{process}. This process consists of five phases: Matching of data sets, grouping the data for job types, analysing the model parameters, generating a blueprint model, and finally injecting the parameters into this model. Note, only a part of the data is used to calibrate the model, the other part is used for its validation.

The starting point of the simulation process is matching the data sets from different data sources as a prerequisite  of the analysis. In our proof of concept we have three data sources: The CERN WMArchive, CMS GridKa job data and GridKa node performance data. These data sets have to be matched because each set contains different information about an executed job, but none contain the complete information about it.

@@

The next step is the generation of the model of GridKa and the injection of the previously extracted performance parameters. Neither step requires manually modeling. At this point the users have a valid and calibrated model of GridKa with its resource and the workload. The properties of this model are described in \Cref{sec:model}.

@@

\Cref{seff} shows the behaviour of the job during the simulation. After a job is scheduled on a resource container, it acquires the specified number of job slots or waits if the node does not have sufficient free slots. Afterwards, the CPU and I/O demands of the job are processed. To avoid one big I/O operation at the start of a job and to make the simulated processing of I/O more continuous, it is modeled using iterations that reflect the number of processed events. Thus, for each iteration there are I/O and CPU resources needed. Rather than simulating each event, we simulate a fraction of the processed events, because the simulation of each event would harm the scalability of the simulation. 
Whilst processing, the job uses the specified number of threads to process the CPU demand. Once the job is finished, it releases its acquired job slots. All of these parameters are specified as a probability function and are extracted from measured performance data as described in \Cref{sec:param}.

@@

At GridKa every type of computing job can run on every node, thus each job type is allocated to each computing node in the allocation model.

@@

We extended Palladio with a \textit{load balancing strategy action}\footurl{https://doi.org/10.5281/zenodo.1464569}, so that we can simulate load balancing strategies.
This load balancing strategy action is contained in a load balancing component.
In the assembly model all job components are connected with the load balancing component, which is the entry point of the system.
This action decides to which node a job should be scheduled, based on the currently free job slots of each working node. The load balancing strategies can be extended to evaluate different strategies.

@@

To obtain a comprehensive view both of the resource usage of CMS jobs executed at a WLCG grid site as well as the resources available there to process them, combining information from multiple data sources is required:
\begin{itemize}
    \item We acquire historical job monitoring information from several global CMS monitoring systems: Job reports from JobMonitoring and Workflow and Data Management framework job reports (FWJR documents) are archived on the \emph{analytix} Hadoop cluster.
    The latter are archived on the cluster as part of the WMArchive project \cite{WMArchiveWorkloadArchive2018}.
    We collect job information using the \emph{CMSSpark} framework by Kuznetsov et al. \cite{cmsspark}.

@@

\begin{figure}[h]
	\centering
	\vspace{-10px}
	\begin{subfigure}{0.48\linewidth}
		\includegraphics[width=\linewidth]{images/validation/throughputs}
	\end{subfigure}
	\begin{subfigure}{0.48\linewidth}
		\includegraphics[width=\linewidth]{images/validation/utilizations}
	\end{subfigure}
	\vspace{-10px}
	\caption{Measured and simulated throughputs and CPU efficiency, job data from May 2018}
	\label{fig:validation}
\end{figure}
\vspace{10px}
\noindent To validate the model and simulation we simulated CMS jobs at GridKa in May 2018.
The job performance data was split into two equal sized calibration and test datasets by randomly sampling reports from the full data set.
For the validation we considered the following metrics: Overall throughput of jobs, shares of the job types, and the overall CPU efficiency of the site.
These metrics describe how well the simulation matches the measured performance data.
Comparisons between simulation results and measurements from the system are shown in \Cref{fig:validation}.
One result is that simulated throughput matches the reference with a difference of about 5.6\%.
Furthermore, the shares of the simulated and measured job types also match closely.

@@

Simulating one week of CMS jobs at GridKa takes about 24 minutes on a laptop.
Thus, it is feasible to simulate several sites at once in a reasonable time, however simulating larger parts of the grid is left as future work.

@@

There are a variety of grid and cloud simulators in existence today, of which SimGrid and CloudSim are two popular examples. 
However, both of them are implemented as libraries and not as standalone simulators. Thus, both are missing graphical model editors which makes the modeling of the computing jobs time-consuming \cite{simgrid,cloudsim}. The support of graphical editors is a big advantage of Palladio and increases its usability and efficiency. Another advantage of Palladio is the separation of the system into different models, which enables one to change one part of the system without modifying others. Furthermore, both lack the support of an integrated visualization tool to display the results.

CACTOS is an approach to cloud infrastructure automation and optimization \cite{cactos}.
One part of the approach is the CACTOS Runtime Toolkit for monitoring and resource management and another part is the CACTOS Prediction Toolkit for evaluation of alternative data center deployment scenarios and resource management algorithms. The Prediction Toolkit is built on Palladio and SimuLizar.
A first case study in the context of chemical computing showed that the Prediction Toolkit has high accuracy, but has not been tested on large systems yet \cite{rapidtesting}. However, due the required Runtime Toolkit CACTOS was not applicable in our use case.

@@

One possible design decision is a more sophisticated load balancing strategy that aims for a better resource utilization of the working nodes by scheduling to it a balanced share of CPU and I/O heavy jobs. To simulate this, a new scheduling strategy must be implemented.

Another design decision would be to experiment with different sizes of pilot jobs, assuming that smaller pilot jobs result in a better utilization due to less wasted CPU capacity in the draining phase of a pilot. For this use case the model must be extended with the explicit modeling of pilots jobs.

Additionally, the effects of integrating cloud resources could be simulated. The model must be extended with the cloud resources and the required network load to access the files from a site. Then a good load balancing strategy for the  integration of cloud resources could be evaluated, such as only scheduling CPU heavy jobs to cloud resources and keeping I/O heavy jobs near the data.

Also the effects of data caches could be simulated. Firstly, several sites must be modeled to simulate their interactions. Secondly, the jobs must be parametrized with the possibility of remote access. Thirdly, the data cache with its hit rate should be modeled. Finally, the performance improvements of the data cache achieved by transferring data from the cache, and not from sites, can be simulated.
\section{Conclusion}
\label{conlusion}

@@

We successfully simulated the CMS computing jobs at GridKa. To enable easy usage of the simulation we implemented an automated model parameter extraction and model creation. Furthermore, we extended Palladio to be able to simulate load balancing strategies. Our first results are promising and suggest that this approach can be used to simulate load balancing strategies accurately. Regarding scalability, we are confident that several computing centers can be simulated in a sufficient time, however this is part of future work.
